Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2022-12-16 unimodular-covers-CombTheory-reviewing.tex
Add files via upload
Rewiewing process for Comb Theory
finished a pass with review Y

A lattice polytope $P\subset \R^d$ has the \emph{integer decomposition property} if for every positive integer $n$, every lattice point $p \in nP\cap \Z^d$ can be written as a sum of $n$ lattice points in $P$. We abbreviate this by saying that ``$P$ is IDP''. Being IDP is interesting in the context of  both enumerative combinatorics (Ehrhart theory) and algebraic geometry (normality of toric varieties). It falls into a hierarchy of several properties each stronger than the previous one; see, e.g., \cite[Section 2.D]{BGbook}, \cite[Sect. 1.2.5]{HPPS-survey}, \cite[p. 2097]{mfo2004}, \cite[p. 2313]{mfo2007}.
Let us here only mention that
\[
P \text{ has a unimodular triangulation}\Rightarrow
P \text{ has a unimodular cover}\Rightarrow
P \text{ is IDP.}
\]
Remember that a \emph{unimodular triangulation} is a triangulation of $P$ into unimodular simplices, and a \emph{unimodular cover} is a collection of unimodular simplices whose union equals $P$.

@@

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$.
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is simply called ``Minkowski summand'' there).

\medskip 
Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes,
with special attention to dimension $3$ (in dimension 2 is it straightforward that ever lattice polygon has unimodular triangulations). 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $P,Q\subset \R^d$ is the lattice polytope
\[
\cayley(P,Q) := \conv(P\times\{0\} \cup Q \times \{1\}) \subset \R^3.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but this does not need $P$ and $Q$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

@@

\begin{definition}
\label{def:circunpara}
Let $T$ be a tetrahedron with vertices $p_1$, $p_2$, $p_3$, and $p_4$. Consider the points $q_i= \frac12 (p_1+p_2+p_3+p_4) - p_i$, $i\in [4]$, and let
\[
C(T)=\conv(p_i,q_i: i\in[4]).
\] 
$C(T)$ is a parallelepiped with facets $\conv(p_i, p_j, q_k, q_l)$ for all choices of $\{i,j,k,l\}=[4]$. We call it the \emph{parallelepiped circumscribed} to $T$.

For each $i \in [4]$, let $T_i=\conv(q_i, p_j, p_k, p_l)$, with $\{i,j,k,l\}=[4]$; we call these $T_i$ the \emph{corner tetrahedra} of $C(T)$. Together with $T$ they triangulate $C(T)$.
\end{definition}

@@

\begin{lemma}
\label{lemma:corner}
Let $T=\conv\{p_1,p_2,p_3,p_4\}$ be an empty lattice tetrahedron that is not unimodular. Let $C(T)$ be the parallelepiped circumscribed to $T$ and let $T_1, T_2,T_3$ and $T_4$ be the corresponding corner tetrahedra in $C(T)$. Then, every $T_i$ contains at least one lattice point different from $\{p_1,\dots,p_4\}$.
\end{lemma}

@@

\begin{lemma}
\label{lemma:3<4}
Let $P$ be a lattice parallelepiped and let $T\subset P$ be a tetrahedron. Then, at least one of the four corner tetrahedra $T_i$ of the circumscribed parallelogram $C(T)$ is fully contained in $P$.
\end{lemma}

@@

We call \emph{band} any region of the form $f^{-1}([\alpha,\beta])$ for some functional $f\in (\R^3)^*$ and closed interval $[\alpha,\beta]\subset \R$.
We claim that any band containing $T$ must contain at least three of the $q_i$s. 
This claim implies that the parallelepiped $P$, which is the intersection of three bands, contains at least one of the $q_i$s and hence it fully contains the corresponding $T_i$.

To prove the claim, suppose that $q_1\not\in B:= f^{-1}([\alpha,\beta])$ for a certain band $B \supset T$. 
Without loss of generality, say $f(q_1)<\alpha$. Then the equalities $q_1+q_2=p_3+p_4$ and $q_1+p_1=q_2+p_2$ respectively give:
\begin{gather}
\label{eq:first}
f(q_2) = f(p_3+p_4-q_1) = f(p_3)+f(p_4)-f(q_1) > 2\alpha-\alpha=\alpha,\\
\label{eq:second}
f(q_2) = f(q_2+p_2-p_1) = f(q_2)+(f(p_2)-f(p_1)) < \alpha + (\beta-\alpha) = \beta,
\end{gather}
so that $q_2 \in B$.

@@

The translation of vector $\frac12 (p_1+p_2-p_3-p_4)$ sends $q_1,q_2,p_3,p_4$ to $p_2,p_1,q_4,q_3$ (in this order). By applying this to inequality \eqref{eq:third}, we obtain
\[
\alpha \le f(p_2) < f(q_i) < f(p_1) \le \beta, \quad\text{ for $i=3,4$},
\]
so that $q_3,q_4 \in B$.
%
This finishes the proof of the claim, and of the lemma.

@@

\begin{corollary}
\label{coro:coverpara}
Let $T$ be an empty lattice tetrahedron contained in a lattice parallelepiped $P$. Then, $T$ can be covered by unimodular tetrahedra contained in $P$.
\end{corollary}

\begin{proof}
We proceed by induction on the (normalized) volume of $T$, which is a positive integer. If this volume equals $1$ then $T$ is unimodular and there is nothing to prove, so we assume $T$ is not unimodular. Let $p_1, p_2, p_3, p_4$ denote the vertices of $T$.

@@

Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$ because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis they can be covered unimodularly.
\end{proof}

@@

Let us say that a lattice $3$-polytope $P$ \emph{has the circumscribed parallelepiped property} if it satisfies the conclusion of \Cref{lemma:3<4}:  ``for every empty tetrahedron $T$ contained in $P$ at least one of the four corner tetrahedra in $C(T)$ is 
contained in $P$''. 
If this holds then $P$ has a unimodular cover, since then the proofs of \Cref{coro:coverpara} and \Cref{thm:parallelepipeds} work for $P$.
In turn, our proof that parallelepipeds have the property (\Cref{lemma:3<4}) is based on the fact that they have only three (pairs of) normal vectors. 
In the following two examples we show a smooth $3$-polytope and two $3$-polytope with four normal vectors that do not have the property. The latter are not IDP:

\begin{example}[A smooth 3-polytope without the circumscribed parallelepiped property]
 Let $P$ be the Cayley embedding of a long horizontal rectangle and a long vertical rectangle. That is,
$P=\conv([0,a]\times [0,1] \times\{0\} \cup [0,1]\times [0,b] \times\{1\})$, for big $a$ and $b$. This is smooth and contains the big empty tetrahedron $T$ with vertices (0,0,0), (a,1,0), (0,0,1), (1,b,1) which occupies most of its volume. In particular, none of the corner tetrahedra of $T$ is contained in $P$.
\end{example}

@@

An affirmative answer to the following question (weaker than the circumscribed parallelepiped property) would still  imply that smooth $3$-polytopes can be unimodularly covered, and hence\Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three is:

@@

\begin{proposition}[see, e.g.~\protect{\cite[Thm.~0.4]{Tsuchiya}}]
\label{prop:mixedIDP}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is mixed IDP.
\end{proposition}

@@

\[
\begin{array}{ccc}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow& \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow& \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow& \text{unimodular prod-simplices in $P + Q$}.
\end{array}
\]
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from a vertex of $T_1$ and from a vertex of $T_2$ form a unimodular basis.

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. A triangulation of $\cayley(P,Q)\subset \R^3$ consists of tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$, which are Cayley sums of a triangle in $P$ and a point in $Q$, or viceversa, are automatically unimodular. The case that we need to study are therefore tetrahedra of type $(2,2)$, which are Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$.
The following lemma, whose proof we postpone to \Cref{sec:the_lemma}, is crucial to understand how to unimodularly cover these tetrahedra. 
We use the following conventions: if $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open segments with endpoints $a,b$. Given a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and  the line spanned by $\vec s$ by $\vecline s$.

@@

Let us denote $\tilde u$, $\tilde p_1$, $\tilde p_2$, $\tilde q_1$, $\tilde q_2$ the points corresponding to $u, p_1, p_2, q_1, q_2$ in $\cayley(P,Q)$.
That is, $\tilde p_i = p\times\{0\}$, $\tilde q_i = p\times\{1\}$, and $\tilde u = u\times\{0\}$.
Observe that the assumption $u\in((p_1, p_2) + \vecline q$ implies that of the segments $[u,q_i]$ crosses the triangle $\conv(p_1,p_2,q_j)$, where $\{i,j\}=\{1,2\}$, see \Cref{fig:flip}.

\begin{figure}[htb]
\includegraphics[scale=.3]{flip}
\caption{$[u,q_2]$ intersects $\conv(p_1,p_2,q_1)$}
\label{fig:flip}
\end{figure}

In turn, this means that the polytope $\conv(\tilde u, \tilde p_1, \tilde p_2, \tilde q_1, \tilde q_2) = \cayley(t,q)$
 has the following two triangulations:
\begin{gather*}
\mathcal T^+:= \left\{ \cayley(p,q), \cayley(t, \{q_i\}) \right\},
\\
\mathcal T^-:= \{ \cayley([p_1,u],q), \cayley([p_2,u],q), \cayley(t, \{q_j\}) \}.
\end{gather*}
The tetrahedra $\cayley(t, \{q_j\})$ and $\cayley(t, \{q_i\})$ are unimodular, which implies that $T=\cayley(p,q)$ has volume equal to the sum of the volumes of $\cayley([p_1,u],q)$ and $\cayley([p_2,u],q)$. In particular, we have covered $T$ by the three tetrahedra in $\mathcal T^-$, which are of smaller volume and hence have unimodular covers by inductive assumption.
\end{proof}

@@

Observe that in the strip $q +\vecline p$, there is a unique lattice point on the line $f_q(x)=-1$;  indeed, since $q$ is primitive, the only way that in the strip there could be two lattice points on $f_q(x)=-1$ is if they were on the boundary of the strip, which would however imply that $p+q$ is a unimodular paralellogram, against our assumptions.
Since translating the polytopes by lattice vectors will not result in any loss of generality, we can assume that $p_1$ is that unique lattice point. That is, $f_q(p_1)=-1$, or equivalently, the triangle $\conv(q_1, q_2, p_1)$ is unimodular. Similarly, the unique lattice point in the strip on the line $f_q(x)=1$ is then $q_1+q_2 -p_1$.

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq 0\}$ and $V_2=\{f_p(x) \geq 0\}$. 
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants. 
See \Cref{fig:setup}.
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

@@

\begin{claim}
The following inequalities hold, 
\begin{align*}
\width_{f_q}(\ell) , 
\width_{f_q}(r) ,
\width_{f_p}(t) ,
\width_{f_p}(b) \geq w.
\end{align*}
Each inequality is strict, unless the segment in question coincides with $p$ or $q$.
\end{claim}

@@

If one of the inequalities, say the one for $\ell$, is not strict, then $\ell$ has one endpoint on each of the boundary lines of $(p + \vecline q)$. Unless $\ell = p$, one of the endpoints of $\ell$ is not an endpoint of $p$, say $l_1 \neq p_1$. Thus the triangle $T=\conv(p_2, p_1, l_1)$ is contained in $P$ and its edge $[p_1, l_1]$ is an integer dilation of $q$. Since $\width_{f_q}(T) =w \geq 2$, $T$ must contain a lattice point in the interior of the strip.
\end{proof}

@@

\giulia{unless b or t are equal to q}

\begin{proof}
Suppose by contradiction that the maximum of $f_q$ on $t$  lies in $V_1$ and that the maximum on $b$ lies in $V_2$.

Then $Q \cap V_2$ is contained in the open strip $\{-1<f_q(x)<w-1\}$, of width $w$. This cannot contain a translated copy of $r$, since $\width_{f_q}(r) \geq w$, see \Cref{fig:claim2}. This is a contradiction, since $P$ is a Minkowski summand of $Q$ and therefore $Q$ must have an edge parallel to $t$.
\end{proof}

@@

\begin{claim}
\label{claim:r}
Assume without loss of generality that $b$ and $t$ either are parallel or their affine spans cross in $V_2$ (if they cross in $V_1$, the same claim can be reworded for the segment $l$). Then, 
\begin{enumerate}
\item The intersection of $Q$ with any line parallel to $p$ in $V_2$ has width w.r.t. $f_q$ strictly smaller than $w$.
\item $f_p(r_2) > f_p(r_1)$, that is, $f_p$ achieves its maximum over $r$ in $H_2$.
\end{enumerate}
\end{claim}

\begin{proof}
Both $t$ and $b$ must intersect $p$, otherwise $p_1$ or $p_2$ are the lattice points we are looking for in $Q$.  Their intersections with $p$ are thus endpoints of a segment of width w.r.t $f_q$ less than $w$, the width of $p$. Since $t$ and $b$ cross in $V_2$, the same is true for any segment parallel to $p$ contained in $Q \cap V_2$.

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption on $p_1$, the values of $g$ on these segments differer by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]
\giulia{changed the +1 to a -1, so that the following claim is true}

>>>>

A lattice polytope $P\subset \R^d$ has the \emph{integer decomposition property} if for every positive integer $n$, every lattice point $p \in nP\cap \Z^d$ can be written as a sum of $n$ lattice points in $P$. We abbreviate this by saying that ``$P$ is IDP''. Being IDP is interesting in the context of  both enumerative combinatorics (Ehrhart theory) and algebraic geometry (projective normality of toric
varieties). It falls into a hierarchy of several properties each stronger than the previous one; see, e.g., \cite[Section 2.D]{BGbook}, \cite[Sect. 1.2.5]{HPPS-survey}, \cite[p. 2097]{mfo2004}, \cite[p. 2313]{mfo2007}.
Let us here only mention that
\[
P \text{ has a unimodular triangulation}\Rightarrow
P \text{ has a unimodular cover}\Rightarrow
P \text{ is IDP.}
\]
Remember that a \emph{unimodular triangulation} is a triangulation of $P$ into unimodular simplices, and a \emph{unimodular cover} is a collection of unimodular simplices whose union equals $P$.

@@

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$. \Giulia{RefY says to add the definition of Minkowski sum here, rather than have the definition in section 3; I think it would break up the flow of the introduction. Perhaps a compromise is to add a comment like, "where + denotes the Minkowski sum defined in Section 3". Then again, the Cayley sum is defined both later in the intro and in section 3. What do you think?}
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Thm.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is called ``$\N$-Minkowski summand'' there).

\medskip 
Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes,
with special attention to dimension $3$ (in dimension 2 is it straightforward that ever lattice polygon has unimodular triangulations). 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds (affine images of $3$-cubes) and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $P,Q\subset \R^d$ is the lattice polytope
\[
\cayley(P,Q) := \conv(P\times\{0\} \cup Q \times \{1\}) \subset \R^{d+1}.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but $P$ or $Q$ do not necessarily need to be fulldimensional. We only require the linear subspaces parallel to them to span $\R^d$.

@@

\begin{definition}
\label{def:circunpara}
Let $T$ be a tetrahedron with vertices $p_1$, $p_2$, $p_3$, and $p_4$. Consider the points $q_i= \frac12 (p_1+p_2+p_3+p_4) - p_i$ for each $i\in \{1,2,3,4\}$, and let
\[
C(T)=\conv(p_i,q_i: i\in\{1,2,3,4\}).
\] 
$C(T)$ is a parallelepiped with facets $\conv(p_i, p_j, q_k, q_l)$ for all choices of $\{i,j,k,l\}=\{1,2,3,4\}$. We call it the \emph{parallelepiped circumscribed} to $T$.

For each $i \in \{1,2,3,4\}$, let $T_i=\conv(q_i, p_j, p_k, p_l)$, with $\{i,j,k,l\}=\{1,2,3,4\}$; we call the $T_i$ \emph{corner tetrahedra} of $C(T)$. Together with $T$ they triangulate $C(T)$.
\end{definition}

@@

Observe that the points $q_i$ need not be lattice points. However, the following lemma shows that we can find lattice points in each corner tetrahedron.
\begin{lemma}
\label{lemma:corner}
Let $T=\conv\{p_1,p_2,p_3,p_4\}$ be an empty lattice tetrahedron that is not unimodular. Let $C(T)$ be the parallelepiped circumscribed to $T$ and let $T_1, T_2,T_3$ and $T_4$ be the corresponding corner tetrahedra in $C(T)$. Then, every $T_i$ contains at least one lattice point different from $\{p_1,\dots,p_4\}$.
\end{lemma}

@@

\begin{lemma}
\label{lemma:3<4}
Let $P$ be a lattice parallelepiped and let $T\subset P$ be a tetrahedron. Then, at least one of the four corner tetrahedra $T_i$ of the circumscribed parallelepiped $C(T)$ is fully contained in $P$.
\end{lemma}

@@

We call \emph{band} any region of the form $f^{-1}([\alpha,\beta])$ for some functional $f\in (\R^3)^*\setminus \{0\}$ and closed interval $[\alpha,\beta]\subset \R$.
We claim that any band containing $T$ must contain at least three of the $q_i$s. 
This claim implies that the parallelepiped $P$, which is the intersection of three bands, contains at least one of the $q_i$s and hence it fully contains the corresponding $T_i$.

To prove the claim, suppose that $q_1\not\in B:= f^{-1}([\alpha,\beta])$ for a certain band $B \supset T$. 
Without loss of generality, say $f(q_1)<\alpha$. Then the equalities $q_1+q_2=p_3+p_4$ and $q_1+p_1=q_2+p_2$ respectively give:
\begin{gather}
\label{eq:first}
f(q_2) = f(p_3+p_4-q_1) = f(p_3)+f(p_4)-f(q_1) > 2\alpha-\alpha=\alpha,\\
\label{eq:second}
f(q_2) = f(q_1+p_1-p_2) = f(q_1)+(f(p_1)-f(p_2)) < \alpha + (\beta-\alpha) = \beta,
\end{gather}
so that $q_2 \in B$.

@@

The translation by the vector $\frac12 (p_1+p_2-p_3-p_4)$ sends $q_1,q_2,p_3,p_4$ to $p_2,p_1,q_4,q_3$ (in this order). By applying this to inequality \eqref{eq:third}, we obtain
\[
\alpha \le f(p_2) < f(q_i) < f(p_1) \le \beta, \quad\text{ for $i=3,4$},
\]
so that $q_3,q_4 \in B$.
%
This concludes the proof of the claim, and of the lemma.

@@

\begin{corollary}
\label{coro:coverpara}
Let $T$ be an empty lattice tetrahedron contained in a lattice parallelepiped $P$. Then, $T$ can be covered by unimodular tetrahedra contained in $P$.
\end{corollary}
\giulia{I would ignore comment Y.16, where if I understand correctly a different proof idea is suggested...}
\begin{proof}
We proceed by induction on the (normalized) volume of $T$, which is a positive integer. If this volume equals $1$ then $T$ is unimodular and there is nothing to prove, so we assume $T$ is not unimodular. Let $p_1, p_2, p_3, p_4$ denote the vertices of $T$.

@@

Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$ because, for each $i$, $p_i$ is the unique point of $C(T)$ maximizing the distance to the opposite facet $\conv(p_j,p_k,p_l)$ of $T$. Thus, $S_1$, $S_2$ and $S_3$ cover $T$ and have volume strictly smaller than $T$. The $S_i$ may not be empty, but we can triangulate them into empty tetrahedra, which by inductive hypothesis can be covered unimodularly.
\end{proof}

@@

Let us say that a lattice $3$-polytope $P$ \emph{has the circumscribed parallelepiped property} if it satisfies the conclusion of \Cref{lemma:3<4}:  ``for every empty tetrahedron $T$ contained in $P$ at least one of the four corner tetrahedra in $C(T)$ is 
contained in $P$''. 
If this holds then $P$ has a unimodular cover, since then the proofs of \Cref{coro:coverpara} and \Cref{thm:parallelepipeds} work for $P$.
In turn, our proof that parallelepipeds have the property (\Cref{lemma:3<4}) is based on the fact that they have only three (pairs of) normal vectors. 
In the following two examples we show a smooth $3$-polytope and two $3$-polytopes with four normal vectors that do not have the property. The latter are not IDP:

\begin{example}[A smooth 3-polytope without the circumscribed parallelepiped property]
 Let $P$ be the Cayley embedding of a long horizontal rectangle and a long vertical rectangle. That is,
$P=\conv([0,a]\times [0,1] \times\{0\} \cup [0,1]\times [0,b] \times\{1\})$, for big $a$ and $b$. This is smooth and contains the big empty tetrahedron $T$ with vertices (0,0,0), (a,1,0), (0,0,1), (1,b,1) which occupies most of its volume. In particular, none of the corner tetrahedra of $T$ is contained in $P$. Another way to see this is to write explicitly the remaining vertices $q_i$ of the circumscribed parallelepiped, which are $(\frac{a+1}{2}, \frac{b+1}{2}, 1), (\frac{a+1}{2}, \frac{b+1}{2}, 0), (\frac{1-a}{2}, \frac{b-1}{2}, 1), (\frac{a-1}{2}, \frac{1-b}{2}, 1)$. None of these points are contained in $P$, and therefore none of the corner tetrahedra are either.
\end{example}\giulia{Added the last part to answer comment Y.19}

@@

An affirmative answer to the following question (weaker than the circumscribed parallelepiped property) would still  imply that smooth $3$-polytopes can be unimodularly covered, and hence~\Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three is:

@@

\begin{proposition}[see, e.g.~\protect{\cite[Thm.~0.4]{Tsuchiya}}]
\label{prop:mixedIDP}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is mixed IDP.
\end{proposition} \giulia{We just called the pair IDP in the introduction... Which notation do we prefer?}

@@

\[
\begin{array}{ccc}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow& \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow& \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow& \text{unimodular prod-simplices in $P + Q$}.
\end{array}
\]
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from any vertex of $T_1$ and from any vertex of $T_2$ form a unimodular basis.

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. A triangulation of $\cayley(P,Q)\subset \R^3$ consists of tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$, which are Cayley sums of a triangle in $P$ and a point in $Q$, or viceversa, are automatically unimodular. The case that we need to study are therefore tetrahedra of type $(2,2)$, which are Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$. These correspond to prod-simplices of two segments in $P+Q$, which are parallelograms.
The following lemma, whose proof we postpone to \Cref{sec:the_lemma}, is crucial to understand how to unimodularly cover these tetrahedra. 
We use the following conventions: if $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open line segments with endpoints $a,b$. Given a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and  the line spanned by $\vec s$ by $\vecline s$.

@@

Let us denote $\tilde u$, $\tilde p_1$, $\tilde p_2$, $\tilde q_1$, $\tilde q_2$ the points corresponding to $u, p_1, p_2, q_1, q_2$ in $\cayley(P,Q)$.
That is, $\tilde p_i = p_i\times\{1\}$, $\tilde q_i = 	q_i\times\{0\}$, and $\tilde u = u\times\{1\}$.\giulia{to match the figure, P is now at height 1 instead of 0}
Observe that the assumption $u\in((p_1, p_2) + \vecline q$ implies that one of the segments $[u,q_i]$ crosses one of the triangles $\conv(p_1,p_2,q_j)$, where $\{i,j\}=\{1,2\}$, see \Cref{fig:flip}.

\begin{figure}[htb]
\includegraphics[scale=.3]{flip}
\caption{$[\tilde{u},\tilde{q}_2]$ intersects $\conv(\tilde{p}_1,\tilde{p}_2,\tilde{q}_1)$}
\label{fig:flip}
\end{figure}\giulia{added tildes to the points}

In turn, this means that the polytope $\conv(\tilde u, \tilde p_1, \tilde p_2, \tilde q_1, \tilde q_2) = \cayley(t,q)$
 has the following two triangulations:
\begin{gather*}
\mathcal T^+:= \left\{ \cayley(p,q), \cayley(t, \{q_j\}) \right\},
\\
\mathcal T^-:= \{ \cayley([p_1,u],q), \cayley([p_2,u],q), \cayley(t, \{q_i\}) \}.
\end{gather*}
The tetrahedra $\cayley(t, \{q_j\})$ and $\cayley(t, \{q_i\})$ are unimodular, which implies that $T=\cayley(p,q)$ has volume equal to the sum of the volumes of $\cayley([p_1,u],q)$ and $\cayley([p_2,u],q)$. In particular, we have covered $T$ by the three tetrahedra in $\mathcal T^-$, which are of smaller volume and hence have unimodular covers by induction hypothesis.
\end{proof}

@@

We can assume that $q_1$ is the origin, and thus $f_q(q)=0$. Observe that in the strip $q +\vecline p$, there is a unique lattice point on the line $f_q(x)=-1$;  indeed, since $q$ is primitive, the only way that in the strip there could be two lattice points on $f_q(x)=-1$ is if they were on the boundary of the strip, which would however imply that $p+q$ is a unimodular paralellogram, against our assumptions.
Since translating the polytopes independently by lattice vectors will not result in any loss of generality, we can assume that $p_1$ is that unique lattice point. That is, $f_q(p_1)=-1$, or equivalently, the triangle $\conv(q_1, q_2, p_1)$ is unimodular. Similarly, the unique lattice point in the strip on the line $f_q(x)=1$ is then $q_1+q_2 -p_1$.

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq f_p(p)\}$ and $V_2=\{f_p(x) \geq f_p(p)\}$. 
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants (see \Cref{fig:setup}). This is always possible via an affine transformation. \Giulia{I'm not sure this last sentence addition is necessary, but I thought it would help the ref clear up confusion (Y.31) about p being a vertical segment-sure, this isn't possible with a unimod transformation, but we don't need to fix the lattice...}
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

@@

\begin{claim}\label{claim:width}
The following inequalities hold, 
\begin{align*}
\width_{f_q}(\ell) , 
\width_{f_q}(r) ,
\width_{f_p}(t) ,
\width_{f_p}(b) \geq w.
\end{align*}
Each inequality is strict, unless the segment in question coincides with $p$ or $q$.
\end{claim}

@@

If one of the inequalities, say the one for $\ell$, is not strict, then $\ell$ has one endpoint on each of the boundary lines of $(p + \vecline q)$. Unless $\ell = p$, one of the endpoints of $\ell$ is not an endpoint of $p$, say $l_1 \neq p_1$. Thus the triangle $T=\conv(p_2, p_1, l_1)$ is contained in $P$ and its edge $[p_1, l_1]$ is an integer dilation of $q$.

Since $T$ contains $p$ and a copy of $q$, its area (normalized to a fundamental domain) is $w/2 \geq 1$, and by Pick's theorem it must contain a lattice point other than its vertices. Since $p$ and $q$ are primitive, this lattice point must lie in the interior of the strip.\giulia{added an explanation here, according to the suggestion Y.32}
%Since $\width_{f_q}(T) =w \geq 2$, $T$ must contain a lattice point in the interior of the strip.
\end{proof}

@@

\begin{proof}
Were $f_q(b_2-b_1)$ or $f_q(t_2 - t_1)$ zero, that is, $b$ or $t$ horizontal, it would have to contain a translate of $q$ whose endpoints are necessarily on the boundary of the strip. The left (or right) endpoint of this translate and of $q$ are lattice points lying on a line parallel to $p$, and since both this translate of $q$ and $q$ cross the segment $p$, they force $p$ to contain a lattice point in the interior, a contradiction to $p$ being primitive.

Suppose by contradiction that the maximum of $f_q$ on $t$  lies in $V_1$ and that the maximum on $b$ lies in $V_2$. Observe that both $t$ and $b$ must intersect $p$, otherwise $p_1$ or $p_2$ are the lattice points we are looking for in $Q$.
%
Then $Q \cap V_2$ is contained in the open strip $\{-1<f_q(x)<w-1\}$, of width $w$. This cannot contain a translated copy of $r$, since $\width_{f_q}(r) \geq w$, see \Cref{fig:claim2}. This is a contradiction, since $P$ is a weak Minkowski summand of $Q$ and therefore $Q$ must have an edge parallel to $r$.
\end{proof}

@@

\begin{claim}
\label{claim:r}
Assume without loss of generality that $b$ and $t$ either are parallel or their affine spans cross in $V_2$ (if they cross in $V_1$, the same claim can be reworded for $V_1$ and $l$). Then, 
\begin{enumerate}
\item The intersection of $Q$ with any line parallel to $p$ in $V_2$ has width with respect to $f_q$ strictly smaller than $w$.
\item $f_p(r_2) > f_p(r_1)$, that is, $f_p$ achieves its maximum over $r$ in $H_2$.
\end{enumerate}
\end{claim}

\begin{proof}
Both $t$ and $b$ must intersect $p$, as seen in the proof of \Cref{claim:b_and_t}.  Their intersections with $p$ are thus endpoints of a segment of width with respect to $f_q$ less than $w$, the width of $p$. Since $t$ and $b$ cross in $V_2$, the same is true for any segment parallel to $p$ contained in $Q \cap V_2$.

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption on $p_1$, the values of $g$ on these segments differer by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]
\giulia{changed the +1 to a -1, so that the inequalities in the following claim are the right way around--we had flipped everything}
